---
title: 'Class06: R Functions'
author: "Arsam Rostami"
date: "2023-04-21"
output: gfm
---

In this class we will develop our own **R function** to calculate the average grade in a fictional class.

we will start with a simplified version of the problem, Just calculating the average grade of one student.

# Grades of Students

An example model is given using three students with different scores obtained for 8 homeworks to achieve `mean ()` and lowest score dropped through created **function**.

```{r}
# Example input vectors to start with

student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)

```

We are going to start by calculating the [**average scores**]{.underline} of the homework using `mean()`.

```{r}
# To find the average grade of student1 without dropping the lowest score

mean(student1)
```

To get the minimum score of something we can use `which.min` giving the placement of the lowest score.

```{r}
# using which.min to find the placement of the lowest score at the 8th postion

student1
which.min(student1)
```

[**FIRST**]{.underline}, we can do the average of the first 7 using `[1:7]` homework scores:

```{r}
# One way to create a code using [1:7] to calculate only first 7 scores

mean(student1[1:7])
```

[SECOND]{.underline} way to do this:

```{r}
# using minus for the placement score will not count the 8th score

student1[1:7]
student1[-8]
```

### HOWEVER...

[THIRD]{.underline} way to drop the lowest score is going to be more efficient and not reliant of the score position:

```{r}
#use -which.min to drop the lowest score and set it to student1_drop_lowest

student1_drop_lowest = student1[-which.min(student1)]
student1_drop_lowest
```

I can get the mean of the homework scores by the dropping the lowest scores by doing `mean(student1_drop_lowest)`

```{r}
# Use the mean() on student1_drop_lowest

mean(student1_drop_lowest)
```

We have our first working snippet code!

### Now lets try to do this for student2:

```{r}
# use -which.min and set the score to student2_drop_lowest

student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student2_drop_lowest = student2[-which.min(student2)]
student2_drop_lowest

```

[***Note***]{.underline}**:** this code only dropped the numerical value for the minimum

There is a way to calculate the mean dropping missing values **(or NA).** Use `na.rm=TRUE`

```{r}
#since there are more NA use na.rm

student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
mean(student2, na.rm=TRUE)
```

## However, for student 3...

```{r}
#set na.rm to TRUE and take the mean of student3

student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
mean(student3, na.rm = TRUE)
```

We want to know the positions of the **NA's**. So, for the Student2 we can use the following.

```{r}
# Use which with is.na to identify the placement of NA

student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
which(is.na(student2))
```

[**NOTE**]{.underline}**:** these are logical vectors giving the potion of the NA

For student 3:

```{r}
# Use which with is.na to identify the placement of NA

student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
which(is.na(student3))
```

For considering the missing values we can mask the NA values with zeros.

```{r}
# Mask all the NA values with zeros setting is.na value assign to 0

student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)

student2[is.na(student2)] <- 0
student2
```

For student 3:

```{r}
# Set similar parameters for student3 using is.na set to 0

student3[is.na(student3)] <- 0
student3
```

Finding the `mean()` for student3:

```{r}
# Find mean student3

mean(student3)
```

This is going to be our final working snippet of code for all the students (with or without NA)

```{r}
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
student3[is.na(student3)] <- 0
student3_drop_lowest = student3[-which.min(student3)]
mean(student3_drop_lowest)
```

## *Let's built an example function now:*

```{r}
x <- c(100,75,50,NA)
x[is.na(x)] <- 0
x_dro_lowest <- x[-which.min(x)]
mean(x_dro_lowest)
```

# FUNCTION grade()

We can write it as function:

```{r}
#' Calculate the average score for a vector
#' of homework scores, dropping the lowest
#' score, and considering NA values as zero
#'
#' @param x A numeric vector of homework
#' scores
#'
#' @return The average value of homework
#' scores
#' @export
#'
#' @examples
#' 
#' student1 <- c('100,'50', NA)
#' grade(student)
#' 
grade <- function(x){
   # we are masking the values from NA to 0
   x[is.na(x)] <- 0
   # Drop the lowest score
   x_dro_lowest <- x[-which.min(x)]
   mean(x_dro_lowest)
}
```

Let's apply the `function (x)` that is set to `grade` and now w can calculate each students [*average grade with lowest score dropped*]{.underline}

```{r}
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)

grade(student1)
grade(student2)
grade(student3)
```

# Gradebook From URL

Lets apply our **function** to a gradebook from this `URL`: <https://tinyurl.com/gradeinput>

```{r}
URL <- 'https://tinyurl.com/gradeinput'
gradebook <- read.csv(URL, row.names = 1)

```

Lets apply my fucntion `grade` to the gradebook using `apply` and running it by `rows` using `MARGIN=1`

#### **Q1.** Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single score.

```{r}
# Use apply() function on gradebook to identify grade

apply(gradebook, 1, grade)
```

#### **Q2.** Using your grade() function and the supplied gradebook, Who is the top scoring student overall in the gradebook?

```{r}
# Use max with apply on the gradebook to identify which score is the highest

max(apply(gradebook, 1, grade))
```

the maximum score is 94.5.

```{r}
# Use which.max to identify the student with highest score

which.max(apply(gradebook, 1, grade))
```

The student who achieved the highest score was `student-18`

#### **Q3.** From your analysis of the gradebook, which homework was toughest on students (i.e. obtained the lowest scores overall?

First we are going to mask the NA

```{r}
# Mask NA with is.na set to zero

gradebook[is.na(gradebook)] <- 0

```

**NOTE:** This the value without the NA considered as a numeric value

Now, we apply the `mean` function to the gradebook.

```{r}
# Use apply function on gradebook

apply(gradebook, 2, mean)
```

The toughest homework will be hw2 considering the mean, and considering missing homework considered as zero

Maybe having zero values for the NA values are not a good representation model to indicate the difficulty of the homework

One thing we can do is to remove the missing values.

```{r}
gradebook <- read.csv(URL, row.names = 1)
apply(gradebook, 2, mean, na.rm = TRUE)
```

Now based on the turned in assignments, and reconsidering the missing values (NA) as outliers, the toughest homework will be hw3 (according to the mean).

If we use the `median` instead of the mean as a measure of overall score...

```{r}
# Set NA prameters na.rm = TRUE and apply function
apply(gradebook, 2, median, na.rm = TRUE)
```

If we use some plots...

```{r}
# create a bxplot based on gradebook URL
boxplot(gradebook, col = "pink")
```

#### **Q4.** From your analysis of the gradebook, which homework was most predictive of overall score (i.e. highest correlation with average grade score)?

If we want to calculate the correlation of grades, we first need the `overall_grades`

```{r}
# set the apply() parameters to overall_grades
overall_grades = apply(gradebook, 1, grade)
```

```{r}
# correlation value for homework1

cor(gradebook$hw1, overall_grades)
```

```{r}
# correlation value for all homeworks having 2 values

gradebook[is.na(gradebook)] <- 0
apply(gradebook, 2, cor, y= overall_grades)
```

The maximum value is...

```{r}
# Use which.max with apply key to identify MAX value for hw?

which.max(apply(gradebook, 2, cor, y= overall_grades))
```

Based on the `which.max` value hw5 was the highest correlation score to the `mean`
